Light inextensible strings under tension in the Schwarzschild geometry

### 2019-02-13 string_ajp.tex
typo
more rewording

\section{Introduction}
\label{intro}
Light inextensible strings are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
expressions for the strength of gravity at the event horizon of a
black hole gives $L\lesssim TGM/\left(\mu c^2\right)$.  For the string
to behave as though light on lengthscales comparable with the
Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
criterion would violate the dominant energy condition (and in this
context we note that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
light inextensible string is an interesting and informative limit: its
analysis interprets the concept ``straight line'' and suggests a
possible way of manipulating a black hole.  The Schwarzschild metric

@@

Solutions to equation~\ref{rdashdash} fall into two distinct classes:
those with $r>1$ at all points, and those which attain $r=1$, in which
case the string is tangential to the event horizon, in general.
Numerical methods show that if the closest approach is greater than a
critical value of~$\simeq 1.076109317(1)$, the string does not
self-intersect (Figure~\ref{closest_approach_non_self_intersecting});
if the closest approach is less than this value, the then string
crosses itself at~$\phi=\pi/2$
(Figure~\ref{closest_approach_self_intersecting}).  The case where the
string is tangent to the event horizon at~$r=1,\phi=0$ is shown in
Figure~\ref{fixed_EH_intersection}.

\begin{figure}[h!] % Created by "closest_approach.R"
\centering
\includegraphics[width=170mm]{closest_approach_nonselfintersecting.pdf}
\caption{Non-self intersecting strings arranged by distance~$d$ of
  closest approach to the event horizon, occurring tangentially
  at~$\phi=0$}
\label{closest_approach_non_self_intersecting}
\end{figure}

\begin{figure}[h!] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings arranged by closest approach to the
  event horizon which occurs at $\phi=0$.  
  shown for $\phi\geq 0$}
\label{closest_approach_self_intersecting}
\end{figure}

\begin{figure}[h!] % created by "one_free_end_fixed_EH_intersection.R"
\centering
\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
\caption{Strings tangential to event horizon}
\label{fixed_EH_intersection}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics{curvature_switch.pdf} % created by "radius_of_curvature_switch.R"
\caption{Strings arranged by point of curvature switch, which occurs
  at~$\phi=0$.  Colour changes from black to red at points of
  infinite~$R$; black indicates convex to the origin and red indicates
  concave regions}
\label{curvature_switch}
\end{figure}

@@

Calculations are performed using the R programming language; source
code for the figures is available at
\\
\\
\url{https://github.com/RobinHankin/string.git}

>>>>

\label{intro}
Light inextensible strings are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
expressions for the strength of gravity at the event horizon of a
black hole gives $L\lesssim TGM/\left(\mu c^2\right)$.  For the string
to behave as though light on lengthscales comparable with the
Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
criterion would violate the dominant energy condition (and in this
context we note that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
light inextensible string is an interesting and informative limit: its
analysis interprets the concept ``straight line'' and suggests a
possible way of manipulating a black hole.  The Schwarzschild metric

@@

Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  Solutions
fall into two distinct classes: those with $r>1$ at all points, and
those which attain $r=1$, in which case the string is tangential to
the event horizon, unless it is radial.

Taking strings for which $r>1$ first,
figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings.  A bisection
technique reveals that if the closest approach is greater than a
critical value of~$\simeq 1.076109317(1)$, the string does not
self-intersect; if the closest approach is less than this value, then
the string crosses itself at~$\phi=\pi/2$.  As discussed above, if
$\left(r'\right)^2=2r(r-1)$, the string has a point of inflection at
which point the radius of curvature becomes infinite.
Figure~\ref{curvature_switch} shows non-self intersecting strings
rotated so that the point of inflection occurs at~$\phi=0$; we see
that the strings have a finite length in which convex to the origin
and two semi-infinite stretches in which the curvature goes the other
way.

The case where the string is tangent to the event horizon at~$\phi=0$,
including a radial string, is shown in
Figure~\ref{fixed_EH_intersection}.

\begin{figure}[p] % Created by "closest_approach.R"
\centering
\includegraphics[width=170mm]{closest_approach_nonselfintersecting.pdf}
\caption{Non-self intersecting strings arranged by distance~$d$ of
  closest approach to the event horizon, occurring tangentially
  at~$\phi=0$}
\label{closest_approach_non_self_intersecting}
\end{figure}

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings arranged by closest approach to the
  event horizon which occurs at $\phi=0$.  
  shown for $\phi\geq 0$}
\label{closest_approach_self_intersecting}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics{curvature_switch.pdf} % created by "radius_of_curvature_switch.R"
\caption{Strings arranged by point of curvature switch, which occurs
  at~$\phi=0$.  Colour changes from black to red at points of
  infinite~$R$; black indicates convex to the origin and red indicates
  concave regions}
\label{curvature_switch}
\end{figure}

\begin{figure}[p] % created by "one_free_end_fixed_EH_intersection.R"
\centering
\includegraphics[width=170mm]{one_free_end_fixed_EH_intersection.pdf}
\caption{Strings tangential to event horizon}
\label{fixed_EH_intersection}
\end{figure}

@@

Calculations are performed using the R programming language~\citep{rcore2018};
source code for the figures is available at
\\
\\
\url{https://github.com/RobinHankin/string.git}
